Approximation as Differentiation

### 2025-05-24 main/formal-structures.tex
End of 2.1/beginning of 2.2 currently a bit noddy..

Our approach to a denotational account of \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vákár22,nunes2023}. Before we go into details, we present a high-level overview of this approach and how we have applied it to \GPS.

The fundamental approach in CHAD is to consider Grothendieck constructions on indexed categories $T : \cat{C}^\op \to \Cat$. An object of the Grothendieck construction $\int T$~, described in more detail in \secref{Grothendieck}, is a pair $(X \in \cat{C}, \partial X \in T(X))$, which we read as pairing a space $X$ of ``points'' from $\cat{C}$ with an associated bundle of tangent spaces. The maps from the ``base'' category $\cat{C}$ are used to interpret the programs we are modelling. The maps in the indexed category are the maps of tangents or approximations, being derivatives and Galois connections respectively.

@@

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$~(\secref{Fam}). These settings are adequate for interpreting first-order programs, but neither is Cartesian closed. To interpret higher-order programs, the picture is more complicated. As is well known, $\Fam(\cat{X})$ for any category $\cat{X}$ is the free coproduct completion of $\cat{X}$~\cite{lawvere63}. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$ (using the Cartesian products from $\Set$). When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}~(\secref{biproducts}), then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

@@

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{higher-order}, but first we flesh out the first-order setting.

@@

The meet-semilattice homomorphisms from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$
on homomorphisms is defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves
meets; the constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets,
provides the unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given
pointwise by $\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$.
Since these two constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet
semilattice with unit $\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and
meet of Galois connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus
$\LatGal$ is enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

Another notable property of $\LatGal$ is that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y
\to Y$, where $X \times Y$ denotes the product of lattices, have both upper and lower adjoints. This allows $X
\times Y$, which we shall hereafter write as $X \biprod Y$, to act as both a product and a coproduct, with
projections $\biproj_X$ and $\biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we will write as $0$, is the
unit for $\biprod$. Such a structure $(\biprod, 0)$, where $\biprod$ acts as both a product and a coproduct,
is called a \emph{biproduct} (\secref{biproducts}).

$\LatGal$ is not Cartesian closed; the set of Galois connections $X \to Y$ does not in general form a bounded
lattice, for a start. In fact if $\cat{C}$ has biproducts $(\biprod, 0)$ and is Cartesian closed, then every
morphism in $\cat{C}$ is trivial. If $0$ is both initial and terminal, and therefore $0 \times X \iso X$, then
$\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X \Rightarrow Y} \iso 1$ and so any morphism $f: X \to Y$ must be
equal to the composite $X \to 0 \to Y$ of the unique morphisms in and out of $0$. The biproduct structure will
prove crucial for establishing a Cartesian closed setting for interpreting higher-order programs.

>>>>

Our approach to a denotational account of \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vákár22,nunes2023}. Before we go into details, we present a high-level overview of their approach and how we have applied it to \GPS.

The fundamental approach in CHAD is to consider Grothendieck constructions on indexed categories $T : \cat{C}^\op \to \Cat$. An object of the Grothendieck construction $\int T$, described in more detail in \secref{Grothendieck}, is a pair $(X \in \cat{C}, \partial X \in T(X))$, which we read as pairing a space $X$ of ``points'' from $\cat{C}$ with an associated bundle of tangent spaces. The maps from the ``base'' category $\cat{C}$ are used to interpret the programs we are modelling. The maps in the indexed category are the maps of tangents or approximations, being derivatives and Galois connections respectively.

@@

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$~(\secref{Fam}). These settings are adequate for interpreting first-order programs, but neither is Cartesian closed. The picture for higher-order programs is more complicated. As is well known, $\Fam(\cat{X})$ for any category $\cat{X}$ is the free coproduct completion of $\cat{X}$~\cite{lawvere63}. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$, using the Cartesian products from $\Set$. When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}~(\secref{biproducts}), then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

@@

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{higher-order} but first flesh out the first-order setting.

@@

The meet-preserving maps from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$ on such
maps is defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves meets; the
constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets, provides the
unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by
$\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two
constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit
$\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus $\LatGal$ is
enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

$\LatGal$ is not Cartesian closed, and so we cannot use it directly to interpret higher-order programs. In
fact Cartesian closure is incompatible with another notable property of $\LatGal$, which follows from the fact
that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the
product of lattices, have both upper and lower adjoints. This allows $X \times Y$, which we shall hereafter
write as $X \biprod Y$, to act as both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$
and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. Although the biproduct
structure rules out Cartesian closure, it will prove crucial for establishing a Cartesian closed setting in
\secref{higher-order}; first we develop a suitable categorical setting for first-order programs.
